Classical Planning Algorithms ¡ on the Atari Video Games: Preliminary Results

### 2014-11-19 workshop-paper/paper.tex
Submitted Workshop version

\author{Nir, Miquel, H}
%AAAI Press\\
%Association for the Advancement of Artificial Intelligence\\
%2275 East Bayshore Road, Suite 160\\
% Palo Alto, California 94303\\}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013) 
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however, 
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given, 
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually  
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,  
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions, 
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge 
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, 
.., $\IW(k)$ where  $\IW(i)$ is  
a breadth-first search in which a state is  pruned when it is not the first state in the search  to make true some set of i atoms or less.
Preliminary empirical results suggest that the performance of \IW\ with the  $k$ parameter fixed to $1$, 
i.e., $\IW(1)$, is much better than random or breadth-first search, and approaches 
the performance of \UCT. A simple best-first variation  of \IW,  that combine exploration and exploitation,
is evaluated as well. 
\end{quote}
\end{abstract}

@@

The Arcade Learning Environment (ALE)  provides a challenging  platform for evaluating general, domain-independent AI planners and 
learners through a convenient interface to hundreds of Atari 2600 games \cite{bellemare:jair2013}. 
Results have been reported so far for basic planning algorithms like breadth-first 
search and \UCT\ \cite{bellemare:jair2013}, and for  reinforcement learning
and evolutionary  methods \cite{bellemare:jair2013,deep-mind-atari,risto-texas:atari-games}.
The empirical results are impressive in some cases, yet a lot remains to be done, as no method  approaches   the 
performance of  human players across a broad range of games.

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects on the state, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners \cite{geffner:book}.
This is because there is  no compact PDDL-like encoding of the domain, and more 
importantly,  the goal to be achieved in each game  is not given. Indeed, there are 
often no goals  but rewards $r(a,s)$ that  depend on the action $a$ and 
the state $s$ in  which  the action is performed, and these rewards are not known a priori 
either.\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}
Thus no model of the  goals  or the  rewards can be used to bias the search, and hence, 
standard heuristic search and classical planning methods can't be used.

@@

The Iterative Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem 
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the 
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to 
represent the value $x_j$ of variable $X_i$ in the state. 
In addition to the state space, a problem is defined by an initial state $s_0$, 
a set of actions applicable in each state, a 
transition function $f$ such that  $s' = f(a,s)$ is 
the state that results from applying action $a \in A(s)$ to the state $s$, 
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known a priori, yet in that case, the  
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated 
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW consists of a sequence of calls  $\IW(i)$ for $i=0, 1, 2, \ldots$
over a problem $P$ until a termination condition is reached. 
The procedure $\IW(i)$  is a plain forward-state \emph{breadth-first search} with just one change: right after a state 
$s$ is generated, the state is  pruned if it doesn't pass a simple \emph{novelty test}.
More precisely,

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the 
search in a different way. This structured exploration  has proved to be very 
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition. 
Yet, this changes the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows 
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration 
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows, 
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.  
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening, 
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$ 
\cite{bonet:aij-hsp} on the other (this algorithm is similar 
to one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}), 
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than the heuristic search  
\cite{nir:ecai2012}.

The min $k$ value for which $\IW(k)$ solves a problem  is clearly bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

One way to understand this definition is that the problem  width $w(P)$ is at most  $i$ if  there is a 
``trail of  stepping stones''  $t_0, t_1, \ldots, t_n$ to reach the problem goal such that 
A)~these ``stepping stones''  contain at most $i$ atoms each, B)~each  stepping stone $t_{i+1}$ is 
at distance 1 from the previous one $t_i$ when $t_i$ has been reached following the ``trail'', 
C)~the ``trail''  preserves ``optimality''; i.e., no  tuple $t_i$ can be reached in less than $i$ steps.

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain  
relevant  in  optimization problems as well. Indeed, if a  good path is made of 
states $s_i$ each of which have a low width, \IW\ can be made 
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property
which is actually  not yet implemented in the code that is tested below.

@@

\IW\ is a purely exploration algorithm that does not take into account  the accumulated 
reward for selecting the next states to consider. As a simple, variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure 
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and 
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on 
both queues. The exception are the nodes with no accumulated reward  that are 
placed in the first queue only. We refer to this best-first algorithm 
as \BFS.

@@

\begin{table*}
\centering
\addtolength{\tabcolsep}{-2pt}
{\scriptsize
\begin{tabular}{@{}l|r|rr|rr|rr|rr@{}}
%% table header
\hline\\[-3ex]
\multicolumn{1}{c|}{} & \multicolumn{1}{c|}{Random} & \multicolumn{2}{c|}{\BRFS}& \multicolumn{2}{c|}{$\IW(1)$}& \multicolumn{2}{c|}{\BFS}& \multicolumn{2}{c}{\UCT} \\[.2ex]
\hline\\[-3ex]
%% table body
\multicolumn{1}{c}{Game}  & 
\multicolumn{1}{c|}{Avg. Score} & 
\multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  & 
\multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  & 
\multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  & 
\multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c}{Avg. Time}\\[.2ex]
\hline\\[-3ex]
\Asterix        & 210    & 3900  & 32.24 & 119000  & 17.08 & 56320  & 0.95  & \textbf{296300} (290700) & 10.82\\
\BeamRider     & 448.8 & 519.2 & 43.48 & 3000          & 6.02  & 3804        & 0.92  & \textbf{4320} (6624)   & 42.91\\
\Breakout       & 1    & 2     & 51.56 & 383.6         & 1.37  & 421.2       & 3.95  & \textbf{512.4} (364)   & 26.03\\
\Enduro          & 0   & 0.4  & 40.92 & \textbf{500}   & 38.13 & 292.8      & 13.12 & 306.8 (286)             & 32.14\\
\Freeway         & 0   & 0     & 71.84 & \textbf{29.4} & 15.19 & 19.6       & 4.44  & 1 (0)    & 90.64\\
\Pong           & -20.4 & -20.2 & 41.51 & 16.4         & 18.45 & \textbf{21} & 39.69 & \textbf{21} (21) & 38.37\\
\Qbert           & 185   & 160  & 42.83 & 1545         & 1.82  & 14725       & 1.76  & \textbf{18730} (17343) & 28.61\\
\Seaquest       & 192   & 200   & 40.01 & \textbf{4274}& 14.72 & 4070        & 1.59  & 3406 (5132)      & 26.16\\
\SpaceInvaders & 187    & 115   & 46.11 & \textbf{2670} & 12.1  & 1327       & 2.25  & 2634 (2718)  & 19.66
\end{tabular}
}

 \caption{\small Random baseline, \BRFS, $\IW(1)$, \BFS, \UCT~performance over 5 runs,
   setting the Atari 2600 random seeds to $0, \ldots, 4$, and maximum episode duration of $18,000$
   frames. For every algorithm, minus the trivial random baseline algorithm, we report the average CPU time per call 
   to each action selection algorithm, along with the average score achieved. \UCT~performance
   reported in \cite{bellemare:jair2013} is shown in parenthesis. Numbers
   in bold show best performer in terms of average score.
   }
\label{performance}

@@

We tested  $\IW(1)$ and \BFS\  over $9$ games: those considered in the \emph{training} set
of \cite{bellemare:jair2013} and those discussed in \cite{deep-mind-atari}. 
In Table~\ref{performance} the performance of  $\IW(1)$ and \BFS\ 
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead, 
is compared with that of \UCT, breadth-first search (\BRFS), and random action selection. 
Our experimental setup  follows the one in  \cite{bellemare:jair2013} with two exceptions. First, rather than
considering $10$ episodes of each game, we have considered just $5$ due to time constraints, as the experiments 
for each  game can take up to $40$ hours of CPU time. Second, while \UCT\ is still limited to perform up to $500$ base policy rollouts 
of depth $300$, a maximum of budget of  $100,000$ simulated frames is applied to all the algorithms, \UCT\ included. 
%% Was not clear; hope this is what you meant, if not fix
%%% ($150,000$ frames are to be simulated while performing the lookahead), both $\IW(1)$ and \BFS are limited to
% The other \UCT\ parameters like the exploration constant $C$ and the horizon $m$ are the same ($C=0.1$, $m=300$). 
%% matching the setup used during actual gameplay. %%  (an action is selected every $5$ frames, using a no-op for the remaining $4$). %% Needed?
The average scores reported for \UCT\  \cite{bellemare:jair2013}, without this extra limitation, are included
in  parenthesis for reference. $\IW(1)$ and \BFS\ have been  limited to a planning horizon $m=1500$ where one an action is evaluated every 
$5$ frames. The discount factor used by all the algorithms is $\gamma=0.995$, and all of them  reuse the tree built 
in the previous lookahead by setting the root of the tree to the best child of the root, while 
deleting its siblings and their descendants as in \cite{bellemare:jair2013}.
%%% really need what's below? it's not very clear .. at least to me HG
%% We note that  this may have a different effect on algorithms where all branches have the same number of
%% generated nodes, as is the case of \BRFS, than in $\IW(1)$, \BFS\ and \UCT, where the depth of
%% the branches is uneven.

Table~\ref{performance} shows that $\IW(1)$ and \BFS\ both outperform \BRFS\  and the random baseline.
\BRFS\ rarely collects any reward, as the depth of the \BRFS\ search tree 
results in  a lookahead of $0.3$ seconds ($20$  frames or $4$ nodes). 
On the other hand, both $\IW(1)$ and \BFS\ reach states that are  up to $350$--$1500$ frames deep
($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$ doesn't fully use the number of simulation frames allocated
in many cases, as when  the set of unpruned states becomes  particularly small. 
%% case of \BFS, due to having a novelty of $0$ and the planning horizon being limited to $m$
%%% (*** What is novelty $0$??? Not clear why BFS doesn't take all the budget.

$\IW(1)$ outperforms \UCT\ in $4$ out of $9$ games, 
both in terms of  average scores and average times per call. \BFS\ 
is generally faster than $\IW(1)$ but only outperforms \UCT\ in $2$ of the games (\Freeway\ and
\Seaquest). Compared with the \UCT\ results from  \cite{bellemare:jair2013}, shown in parenthesis, 
both $\IW(1)$ and \BFS\ still outperform \UCT\ in $3$  of the  $9$ games (\Breakout, \Enduro\ and \Freeway),
with \BFS\ and \UCT\ tied in another game, \Pong. 
As mentioned in the previous section, the order in which $\IW(1)$ and \BFS\ generate the children
of an expanded node are randomized, something which has a potential effect on the  novelty measures
of nodes, and hence in the search graph. We found that this randomization helps, in particular 
in \Qbert, where not randomizing results  in $\IW(1)$ being trapped into  loops sometimes.

We also  note that in \Freeway,  both $\IW(1)$ and \BFS\ attain a  performance similar to the 
baseline semi-random algorithm \emph{Perturb} in  \cite{bellemare:jair2013}, 
which outperforms   both \UCT\ and  the reinforcement learning methods. 
Likewise,  $\IW(1)$ outperforms the best learning algorithm in \cite{deep-mind-atari}  
in $4$ out of the  $7$ games considered, and \BFS\ does so in $5$  of the $7$ games.

We do not report the performance of $\IW(2)$ because  in our
preliminary tests and according to the discussion in the previous section,  
it doesn't  appear to improve much on  \BRFS, even if  it results in a lookahead
that is $5$ times deeper, yet still far too shallow  to compete with the other 
planning algorithms.

@@

The notion of width underlying the iterated width algorithm was developed in the context
of classical planning in order to understand why most of the hundreds of existing benchmarks
appear to be relatively simple for current planners, even
though classical planning is PSPACE-complete \cite{bylander:complexity}. A  partial answer is that 
most of these domains have a low width, and hence, can be solved in low polynomial time (by \IW)
when goals contain a single atom.  Benchmark problems with multiple atomic goals tend to be 
easy too, as the goals can often be achieved one at a time \cite{nir:ecai2012}.

In the iterated width algorithm, the key notion is the \emph{novelty measure} of a state in the underlying
breadth-first search. These novelty measures make use of the factored representation of the states for
providing a structure to the search: states that have width 1 are explored first in linear time, 
then states that have  width two are explored in quadratic time, and so on. In classical planning problems 
with atomic goals, this way of organizing the search  pays off both theoretically and practically.

>>>>

\author{Nir Lipovetzky \\
      The University of Melbourne \\
   Melbourne, Australia\\
   {\normalsize\url{@unimelb.edu.au}}
   \And
   Miquel Ramirez \\
   RMIT University \\
   Melbourne, Australia\\
   {\normalsize\url{@rmit.edu.au}}
   \And
   Hector Geffner \\
   ICREA \&  Universitat Pompeu Fabra \\
   Barcelona, SPAIN \\
   {\normalsize\url{@upf.edu}}\thanks{firstname.lastname}
}

%AAAI Press\\
%Association for the Advancement of Artificial Intelligence\\
%2275 East Bayshore Road, Suite 160\\
% Palo Alto, California 94303\\}

@@

\begin{abstract}
\begin{quote}
The Atari  2600 games supported in  the Arcade Learning Environment (Bellemare et. al. 2013)
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some set of i atoms.
Preliminary empirical results suggest that the performance of \IW\ with the  $k$ parameter fixed to $1$,
i.e., $\IW(1)$, is much better than random or breadth-first search, and approaches
the performance of \UCT. A simple best-first variation  of \IW\  that combines exploration and exploitation
is evaluated as well.
\end{quote}
\end{abstract}

@@

The Arcade Learning Environment (ALE)  provides a challenging  platform for evaluating general, domain-independent AI planners and
learners through a convenient interface to hundreds of Atari 2600 games \cite{bellemare:jair2013}.
Results have been reported so far for basic planning algorithms like breadth-first
search and  \UCT,  reinforcement learning algorithms, and evolutionary  methods \cite{bellemare:jair2013,deep-mind-atari,risto-texas:atari-games}.
The empirical results are impressive in some cases, yet a lot remains to be done, as no method  approaches   the
performance of  human players across a broad range of games.

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners \cite{geffner:book}.
This is because there is  no compact PDDL-like encoding of the domain, and more
importantly,  the goal to be achieved in each game  is not given. Indeed, there are
often no goals  but rewards $r(a,s)$ that  depend on the action $a$ and
the state $s$ in  which  the action is performed, and these rewards are not known.\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}
Thus no model of the  goals  or the  rewards can be used to bias the search.

@@

The Iterated Width  (\IW) algorithm has been introduced as a classical planning algorithm
that takes a planning problem as an input, and computes an action sequence that solves the problem
as the output \cite{nir:ecai2012}. The algorithm however
applies to a broader range of problems. We will characterize such
problems  by means of  a finite and discrete set of states (the state space)
that correspond to vectors of size $n$. Namely, the
states are  \emph{structured} or \emph{factored} , and we take each of the locations in the vector
to represent a variable $X_i$, and the value at that vector location to
represent the value $x_j$ of variable $X_i$ in the state.
In addition to the state space, a problem is defined by an initial state $s_0$,
a set of actions applicable in each state, a
transition function $f$ such that  $s' = f(a,s)$ is
the state that results from applying action $a \in A(s)$ to the state $s$,
and rewards $r(a,s)$ represented by real numbers that result from  applying action $a$ in state $s$.
The transition and reward functions do not need to be known a priori, yet in that case, the
state and  reward that results from the application of an action in a state needs to be \emph{observable}.
The task is to compute an action sequence $a_0, \ldots, a_m$ for a large horizon $m$
that generates a state sequence $s_0, \ldots, s_{m+1}$ that  maximizes the accumulated
reward $\sum_{i=0}^{m} r(a_i,s_i)$, or that provides a good approximation.

@@

\IW \ consists of a sequence of calls  $\IW(i)$ for $i=0, 1, 2, \ldots$
over a problem $P$ until a termination condition is reached.
The procedure $\IW(i)$  is a plain forward-state \emph{breadth-first search} with just one change: right after a state
$s$ is generated, the state is  pruned if it doesn't pass a simple \emph{novelty test}.
More precisely,

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search in a different way. This structured exploration  has proved to be very
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows,
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening,
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$
\cite{bonet:aij-hsp} on the other (this algorithm is similar
to the one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}),
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of the Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than a heuristic search algotrithm
\cite{nir:ecai2012}.

The min $k$ value for which $\IW(k)$ solves a problem  is bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

One way to understand this definition is that the problem  width $w(P)$ is at most  $i$ if  there is a
``trail of  stepping stones''  $t_0, t_1, \ldots, t_n$ to reach the problem goal such that
A)~these ``stepping stones''  contain at most $i$ atoms each, B)~each  stepping stone $t_{i+1}$ is
at distance 1 from the previous one $t_i$ when $t_i$ has been reached by following the ``trail'',
C)~the ``trail''  preserves ``optimality''; i.e., no  tuple $t_i$ can be reached in less than $i$ steps.

While this notion of width and the iterated  width algorithms that are based on it  have been designed
for problems where a goal state needs to be reached, the notions remain
relevant  in  optimization problems as well. Indeed, if a  good path is made of
states $s_i$ each of which has a low width, \IW\ can be made
to find such path in low polynomial time for a small value of the $k$ parameter.
Later on we will discuss  a slight  change  required in \IW\ to enforce this property
which is actually  not yet implemented in the code that is tested below.

@@

\IW\ is a purely exploration algorithm that does not take into account  the accumulated
reward for selecting the states to consider. As a simple variant that combines
exploration and exploitation, we evaluated a \emph{best-first search} algorithm
with two queues: one queue  ordered first by  novelty measure
(recall that novelty one means that the state is the first one to make some atom true),
and a second queue ordered by accumulated reward. In one iteration, the best first
search picks up the best node from one queue, and in the second iteration it picks
up the best node from the other queue. This way for combining multiple heuristics
is used in the LAMA planner \cite{richter:lama}, and was introduced in the
planner Fast Downward \cite{helmert:fd}. In addition, we break ties
in the first queue favoring states with largest accumulated reward, and
in the second queue, favoring states with smallest novelty measure. Last, when
a node is expanded, it is removed from the queue, and its children are placed on
both queues. The exception are the nodes with no accumulated reward  that are
placed in the first queue only. We refer to this best-first algorithm
as \BFS.

@@

\begin{table*}
\centering
% \addtolength{\tabcolsep}{-2pt}
% {\scriptsize
{\small
\begin{tabular}{@{}l|r|rr|rr|rr|rr@{}}
%% table header
% \hline\\[-3ex]
%\hline
\multicolumn{1}{c|}{} & \multicolumn{1}{c|}{Random} & \multicolumn{2}{c|}{\BRFS}& \multicolumn{2}{c|}{$\IW(1)$}& \multicolumn{2}{c|}{\BFS}&
   \multicolumn{2}{c}{\UCT}
% \\[.2ex]
\\
% \hline\\\\[-3ex]
\hline
%% table body
\multicolumn{1}{c|}{Game}  &
% \multicolumn{1}{c|}{Avg. Score} &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c|}{Avg. Time}  &
% \multicolumn{1}{c}{Avg. Score} & \multicolumn{1}{c}{Avg. Time}\\[.2ex]
 \multicolumn{1}{c|}{Score} &
 \multicolumn{1}{c}{Score} & \multicolumn{1}{c|}{Time}  &
 \multicolumn{1}{c}{Score} & \multicolumn{1}{c|}{Time}  &
 \multicolumn{1}{c}{Score} & \multicolumn{1}{c|}{Time}  &
 \multicolumn{1}{c}{Score} & \multicolumn{1}{c}{Time}
% \\[.2ex]
\\[.2ex]
% \hline\\
\hline
\Asterix        & 210    & 3900  & 32.24 & 119000  & 17.08 & 56320  & 0.95  & \textbf{296300} (290700) & 10.82\\
\BeamRider     & 448.8 & 519.2 & 43.48 & 3000          & 6.02  & 3804        & 0.92  & \textbf{4320} (6624)   & 42.91\\
\Breakout       & 1    & 2     & 51.56 & 383.6         & 1.37  & 421.2       & 3.95  & \textbf{512.4} (364)   & 26.03\\
\Enduro          & 0   & 0.4  & 40.92 & \textbf{500}   & 38.13 & 292.8      & 13.12 & 306.8 (286)             & 32.14\\
\Freeway         & 0   & 0     & 71.84 & \textbf{29.4} & 15.19 & 19.6       & 4.44  & 1 (0)    & 90.64\\
\Pong           & -20.4 & -20.2 & 41.51 & 16.4         & 18.45 & \textbf{21} & 39.69 & \textbf{21} (21) & 38.37\\
\Qbert           & 185   & 160  & 42.83 & 1545         & 1.82  & 14725       & 1.76  & \textbf{18730} (17343) & 28.61\\
\Seaquest       & 192   & 200   & 40.01 & \textbf{4274}& 14.72 & 4070        & 1.59  & 3406 (5132)      & 26.16\\
\SpaceInvaders & 187    & 115   & 46.11 & \textbf{2670} & 12.1  & 1327       & 2.25  & 2634 (2718)  & 19.66
\end{tabular}
}

 \caption{\small Performance that results from random policy and various  lookahead algorithms in 9 Atari games.
The algorithms  are \BRFS, $\IW(1)$, \BFS, and \UCT. Averages  over 5 runs shown with ``random'' seeds
$0$ to $4$, and maximum episode duration of $18,000$  frames. Every algorithm limited to a budget of
100,000 simulated frames, except for UCT which is limited to 500 rollouts of 300 frames depth instead. Average scores and CPU times per action selection shown.
Performance of UCT results  reported in \cite{bellemare:jair2013} shown in parenthesis.
Numbers  in bold show best performer in terms of average score. }
\label{performance}

@@

We tested  $\IW(1)$ and \BFS\  over $9$ games: those considered in the \emph{training} set
of \cite{bellemare:jair2013} and those discussed in \cite{deep-mind-atari}.
In Table~\ref{performance} the performance of  $\IW(1)$ and \BFS\
% , characterized by the average score obtained over all episodes, the standard deviation
% of the score obtained and time (in seconds) per lookahead,
is compared with that of \UCT, breadth-first search (\BRFS), and random action selection.
Our experimental setup  follows the one in  \cite{bellemare:jair2013} with one exception: rather than
considering $10$ runs per games, we considered just $5$ due to time constraints. Following the setup in  \cite{bellemare:jair2013}, while \UCT\ is limited to perform up to $500$ base policy rollouts
of depth $300$\footnote{In \UCT,  $500$ base policy rollouts
of depth $300$ is equivalent to a budget of $150,000$ simulated frames}, a maximum budget of  $100,000$ simulated frames is applied to all the other algorithms.
%% Was not clear; hope this is what you meant, if not fix
%%% ($150,000$ frames are to be simulated while performing the lookahead), both $\IW(1)$ and \BFS are limited to
% The other \UCT\ parameters like the exploration constant $C$ and the horizon $m$ are the same ($C=0.1$, $m=300$).
%% matching the setup used during actual gameplay. %%  (an action is selected every $5$ frames, using a no-op for the remaining $4$). %% Needed?
The average scores reported for \UCT\  \cite{bellemare:jair2013}, are included
in  parenthesis for reference. $\IW(1)$ and \BFS\ have been  limited to a planning horizon $m=1500$ with an action selected
every $5$ frames. The discount factor used by all the algorithms is $\gamma=0.995$, and all of them  reuse the tree built
in the previous lookahead as in \cite{bellemare:jair2013}, by setting the root of the tree to the best child of the root,
while deleting its siblings and their descendants
%%% really need what's below? it's not very clear .. at least to me HG
%% We note that  this may have a different effect on algorithms where all branches have the same number of
%% generated nodes, as is the case of \BRFS, than in $\IW(1)$, \BFS\ and \UCT, where the depth of
%% the branches is uneven.

Table~\ref{performance} shows that $\IW(1)$ and \BFS\ both outperform \BRFS\  and the random baseline.
\BRFS\ rarely collects any reward, as the depth of the \BRFS\ search tree
results in  a lookahead of $0.3$ seconds ($20$  frames or $4$ nodes deep).
On the other hand, both $\IW(1)$ and \BFS\ reach states that are  up to $350$--$1500$ frames deep
($70$--$260$ nodes or $6$--$22$ seconds), even if $\IW(1)$ doesn't always use all the simulation frames
allocated due to its agressive pruning.
%% case of \BFS, due to having a novelty of $0$ and the planning horizon being limited to $m$
%%% (*** What is novelty $0$??? Not clear why BFS doesn't take all the budget.

Interestingly, simple $\IW(1)$ outperforms \UCT\ in $4$ out of $9$ games,
both in terms of  average scores and average times per call. \BFS\ is
faster than $\IW(1)$ and an order-of-magnitude faster than \UCT\ in general,
and it  outscores \UCT\ in $2$ of the games (\Freeway\ and
\Seaquest). Compared with the \UCT\ results in   \cite{bellemare:jair2013}, shown in parenthesis,
both $\IW(1)$ and \BFS\  outperform \UCT\ in $3$ of the  $9$ games (\Breakout, \Enduro\ and \Freeway),
with \BFS\ and \UCT\ tied in  \Pong. It'd  be very interesting to see the results of
\BFS\ when given more time per action, but we don't have those results yet.
As mentioned in the previous section, the order in which $\IW(1)$ and \BFS\ generate the children
of an expanded node are randomized, something which has a potential effect on the  novelty measures
of nodes, and hence in the search graph. We found that this randomization helps, in particular
in \Qbert, where not randomizing results  in $\IW(1)$ being trapped into  loops sometimes.

We also  note that in \Freeway,  both $\IW(1)$ and \BFS\ attain a score similar to the
baseline semi-random algorithm \emph{Perturb} in  \cite{bellemare:jair2013},
which outperforms   both \UCT\ and  the reinforcement learning methods.
Likewise,  $\IW(1)$ outperforms the best learning algorithm in \cite{deep-mind-atari}
in $4$ out of the  $7$ games considered, and \BFS\ does so in $5$  of
the $7$ games.

We do not report the performance of $\IW(2)$ because  in our
preliminary tests and according to the discussion in the previous section,
it doesn't  appear to improve much on  \BRFS, even if  it results in a lookahead
that is $5$ times deeper,  still far too shallow  to compete with the other
planning algorithms.

@@

The notion of width underlying the iterated width algorithm was developed in the context
of classical planning in order to understand why most of the hundreds of existing benchmarks
appear to be relatively simple for current planners, even
though classical planning is PSPACE-complete \cite{bylander:complexity}. A  partial answer is that
most of these domains have a low width, and hence, can be solved in low polynomial time (by \IW)
when goals contain a single atom.  Benchmark problems with multiple atomic goals tend to be
easy too, as the goals can often be achieved one at a time after a simple goal ordering \cite{nir:ecai2012}.

In the iterated width algorithm, the key notion is the \emph{novelty measure} of a state in the underlying
breadth-first search. These novelty measures make use of the factored representation of the states for
providing a structure to the search: states that have width 1 are explored first in linear time,
then states that have  width 2  are explored in quadratic time, and so on. In classical planning problems
with atomic goals, this way of organizing the search  pays off both theoretically and practically.
